Reduced diversity and increased specialization of Alaska salmon fishermen:

### 2016-10-24 salmon/paper/salmon_paper.Rmd
updates from rich

---
title: 'Reduced diversity and increased specialization of Alaska salmon fishermen:
  effects on variation in individual revenues'
author: "Eric, Sean, Ole, Rich, Milo, Jordan, Jen, Alan, others"
date: ''
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document: default
bibliography: refs.bib
---

@@

On a large scale, individual fishers may experience positive outcomes from diversifying their catches; examples include reduced year-to-year variability in annual revenue and increases in average revenue. Fishers can increase their species diversity by owning a permit that allows multiple species to be targeted, or by owning multiple permits. External forces such as management regulations (limited entry,  catch shares) or market forces (permit or fish prices) may disincentive individuals to hold multiple permits. Thus, individuals with a single permit may increase species diversity within the confines of their permit (longer fishing seasons, broader spatial area). Using a long-term dataset of individual salmon fishers in Alaska, we develop new models for understanding how the effects of diversification have changed through time. These changes allow us to compare patterns across regions and gear, as well as better understand how external forces have changed incentives for diversifying (hatchery production and price trends, EVOS?).

@@

Understanding the effects of species diversification on salmon fishers is particularly important, as the number of salmon specialists (defined as fishers holding a single salmon permit) has increased over the last 30 years as a fraction of the total [@anderson_benefits]. For these limited entry salmon fisheries, general trends that have emerged are that more people are participating since the early 2000s (Fig. A1), median revenue is increasing (across individuals, Fig. A2), and landed species diversity declining over time (averaged across individuals, Fig. A6). Across different salmon fisheries, other than a slight negative effect of diversification on variability (similar to [@anderson_benefits]), there appears to be little correlation between diversity and revenue (Fig. A4). In terrestrial agriculture, there are numerous examples where diversification may affect the mean or CV but not both [@purdy_financial_1997], and the same result may be true for salmon fishers. Further, these impacts of diversification are variable over time.

Five species of salmon are encountered in Alaska waters (Chinook [*Oncorhynchus tshawytscha*], chum [*O. keta*], pink [*O. gorbuscha*], sockeye [*O. nerka*], coho [*O. kisutch*]), and the ability of individual salmon fishers to diversify across species is in part a function of geography and fishing gears. For example, salmon may be caught with purse seine gear (n = 6 areas), drift gillnets (5 areas), and set gillnets (12 areas). Other gears include beach seine and troll gears (though these are generally restricted to Kodiak and Southeast Alaska, respectively). While some of these combinations of gears and areas allow fishers to diversify, there are many examples of permits where the catch is dominated by a single species. Perhaps the best examples include fisheries that target sockeye salmon (e.g. Bristol Bay set and drift gillnet fisheries). For these fisheries, specialization is associated with higher revenue and reduced CV in annual revenue [@anderson_benefits].

A secondary motivation for focusing on the effects of diversification for salmon fishers is to better understand the long term consequences of hatchery salmon production on the revenue and variability in revenue of individuals. Following low returns and declining productivity in the mid 1980s, many hatchery programs in the state rapidly increased hatchery releases [@mahnken_historical_1998; @leber_stock_2008]. Many of these hatchery programs increased production to stabilize the variability of fishers catch (and revenue), though these programs may also have impacts on wild production [@ruggerone_productivity_2015]. The intra- and inter-specific effects of large scale hatchery programs has been debated [@hilborn_review_2001; @hilborn_hatcheries_1992; @wertheimer_comment:_2001; @heard_hatchery], but the effects of hatcheries have largely been focused on ecological interactions of aggregate yield to the fishery [@leber_stock_2008] or economic impacts [boyce_alaska_1993], rather than to variability experienced by individual fishermen. [Other history here?]

@@

In addition to the changing species composition of fish available to be harvested, the benefits (or costs) of diversification are affected by a number of exogenous drivers, including marketing, prices, new processing technology [@knapp_trends]. [Break out as own paragraph+add more examples?] Following price crashes in the early 2000s, many of these factors have been positive for salmon fishermen (increasing trends in prices and catches, Fig. A3, A12). In contrast, the late 1980s - early 1990s experienced high abundance and low prices (in part due to collapses of foreign currencies). Combined with the increased trend in hatchery production and resulting changes in species composition of adult salmon, these external factors may have also altered incentives for diversification. If there's an incentive to substitute high value species (Chinook, sockeye) with more abundant low value species (pink, chum), the analog in the terrestrial farming world would be crop substitution of low-volume high value crops (spices) with high-volume low value crops (cereals) (other examples in literature = cocoa, tobacco).

@@

We obtained fisheries landings and revenue data for all permit holders in Alaska from 1985 - 2014 from the Commercial Fisheries Entry Commission (CFEC). We adjusted for inflation by setting all revenues in 2009 USD. Because we were only interested in single permit fisheries, we adopted a simplified version of the filtering steps described in [@anderson_benefits]. After aggregating individual trip-level fish tickets to annual data, we retained individuals who only fished one of the six salmon permits included in a year (Table 1, Fig. A5). Like [@anderson_benefits] we also worked with a log-differenced dataset to remove non-stationary individuals; to avoid complications of individuals switching strategies, we only included pairs of years for a given person where they did not change strategies.

@@

Hatchery release trends for some regions included in our data have been summarized previously (e.g. [@ward_evaluating] summarized the trends for Prince William Sound). Release patterns have been dominated by pink and chum salmon (in recent years 2012-2014 in PWS, for example, pink and chum salmon represent 80% and 18% of total releases, [@ward_evaluating]) and production has increased since the late 1970s (217 million juvenile pink salmon in PWS in 1985 versus 672 million in 2014). Though our analysis with CFEC data doesn't separate hatchery versus wild fish, as a contrast we can make coarse comparisons across regions and species to examine potential effects of hatchery fish on variability. For example, we can compare the variation in the largely hatchery pink salmon returns (earnings and weight) from Prince William Sound to the largely wild pink salmon returns from Southeast Alaska. Similarly, we can compare the variability in earnings from species dominated by hatchery returns (pink, chum) to those with less hatchery influence (Chinook, sockeye). To make these comparisons, we aggregated the earnings and landings by species and permit, then calculated 3-year rolling CVs. As a second comparison, we examined whether long-term participants in any of the fisheries had experienced a change in variability. Similar to [@kasperski_income_2013] who calculated CVs of vessels before and after significant management events, we identified individuals who fished the same permit for at least three years in the first decade of our data (1985-1995) and at least three years in the last decade of the data (2004-2014).

@@

@@

@@

- In terms of participation, why did people drop out in 2000 in SEAK and Alaska Peninsula, but not in PWS? PWS largely seems like a boutique fishery on the Copper River -- which includes a qualtiy component (Copper River at forefront) and a timing issue relative to the rest of the Gulf of Alaska populations (buffers against market impacts)

@@

- One pontential exogenous driver related to management that can be ruled out as being responsible for increased specialization of salmon fishermen is changes in the timing or length of fishing seasons. Shifting seasons to later in the year may reduce catch of early run fish for example. However, we found no significant trend in either the season length or average season date for any fishery (Fig. A15).

>>>>

---
title: 'Reduced diversity and increased specialization of Alaska salmon fishermen:
  effects on variation in individual revenues'
author: "Eric, Sean, Ole, Rich, Milo, Jordan, Jen, Alan, others"
date: ''
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document: default
bibliography: salmon.bib
---

@@

On a large scale, individual fishers may experience positive outcomes from diversifying their catches; examples include reduced year-to-year variability in annual revenue and increases in average revenue. Fishers can increase their species diversity by owning a permit that allows multiple species to be targeted, or by owning multiple permits. External forces such as management regulations (limited entry,  catch shares), the number of permits available, or market forces (permit or fish prices) may disincentive individuals to hold multiple permits. Thus, individuals holding a single permit may only be able to increase species diversity within the confines of their permit (longer fishing seasons, broader spatial area). Using a long-term dataset of individual salmon fishers in Alaska, we develop new models for understanding how the effects of diversification have changed through time. These changes allow us to compare patterns across regions and gear, as well as better understand how external forces have changed incentives for diversifying (hatchery production, market forces, EVOS?).

@@

Understanding the effects of species diversification on salmon fishers is particularly important, as the number of salmon specialists (defined as fishers holding a single salmon permit) has increased over the last 30 years as a fraction of the total [@anderson_benefits] (figure XX). For these limited entry salmon fisheries, general trends that have emerged are that more people are participating since the early 2000s (Fig. A1), median revenue is increasing (across individuals, Fig. A2), and landed species diversity declining over time (averaged across individuals, Fig. A6). Across different salmon fisheries -- other than a slight negative effect of diversification on variability (similar to [@anderson_benefits]) -- there appears to be little correlation between diversity and revenue (Fig. A4). In terrestrial agriculture, there are numerous examples where diversification may affect the mean or CV of revenue, but not both [@purdy_financial_1997], and the same result may be true for salmon fishers. Further, these impacts of diversification are variable over time.

Five species of salmon are harvested in Alaska waters (Chinook [*Oncorhynchus tshawytscha*], chum [*O. keta*], pink [*O. gorbuscha*], sockeye [*O. nerka*], coho [*O. kisutch*]), and the ability of individual salmon fishers to diversify across species is in part a function of geography and fishing gears. For example, salmon may be caught with purse seine gear (n = 6 areas), drift gillnets (5 areas), and set gillnets (12 areas). Other gears include beach seine and troll gears (though these are generally restricted to Kodiak and Southeast Alaska, respectively). While some of these combinations of gears and areas allow fishers to diversify, there are many examples of permits where the catch is dominated by a single species. Perhaps the best examples include fisheries that target sockeye salmon (e.g. Bristol Bay set and drift gillnet fisheries). For these fisheries, specialization is associated with higher revenue and reduced CV in annual revenue [@anderson_benefits].

A secondary motivation for focusing on the effects of diversification for salmon fishers is to better understand the long term consequences of hatchery salmon production on the revenue and variability in revenue of individuals. The modern era of Alaska’s large-scale hatchery program began in the late 1970s (Stopha 2015, Rich will get a reference), with substantial increases in hatchery production during the 1980s following low returns [@mahnken_historical_1998; @leber_stock_2008-1]. Many of these hatchery programs increased production to stabilize the variability of salmon returns [@hilborn_review_2001], though these programs may also have negative impacts on wild production [@wertheimer_relationship_2005; @ruggerone_productivity_2015]. The intra- and inter-specific effects of large scale hatchery programs has been debated [@hilborn_review_2001; @hilborn_hatcheries_1992; @wertheimer_comment:_2001; @heard_hatchery], but the effects of hatcheries have largely been focused on ecological interactions of aggregate yield to the fishery [@leber_stock_2008-1] or economic impacts [@boyce_alaska_1993], rather than to variability experienced by individual fishermen.

@@

In addition to the changing species composition of fish available to be harvested, the benefits (or costs) of diversification are affected by a number of exogenous drivers, including the influx of farmed salmon, consolidation of salmon buyers, marketing, prices, new processing technology [@knapp_trends]. [Break out as own paragraph+add more examples?] Following price crashes in the 1990s and early 2000s (cite CFEC), many of these factors have been positive for salmon fishermen (increasing trends in prices and catches, Fig. A3, A12). In contrast, the late 1980s - early 1990s experienced high abundance and low prices (in part due to collapses of foreign currencies). Combined with the increased trend in hatchery production and resulting changes in species composition of adult salmon, these external factors may have also altered incentives for diversification. If there's an incentive to substitute high value species (Chinook, sockeye) with more abundant low value species (pink, chum), the analog in the terrestrial farming world would be crop substitution of low-volume high value crops (spices) with high-volume low value crops (cereals) (other examples in literature = cocoa, tobacco).

@@

We obtained fisheries landings and revenue data for all permit holders in Alaska from 1985-2014 from the Commercial Fisheries Entry Commission (CFEC). We adjusted for inflation by setting all revenues in 2009 USD. Because we were only interested in single permit fisheries, we adopted a simplified version of the filtering steps described in [@anderson_benefits]. After aggregating individual trip-level fish tickets to annual data, we retained individuals who only fished one of the six salmon permits included in a year (Table 1, Fig. A5). Like [@anderson_benefits] we also worked with a log-differenced dataset to remove non-stationary individuals; to avoid complications of individuals switching strategies, we only included pairs of years for a given person where they did not change strategies (i.e., didn’t change fisheries or add permits, etc.).

@@

Hatchery release trends for some regions included in our data have been summarized previously (e.g. [@ward_evaluating] summarized the trends for Prince William Sound). Release patterns have been dominated by pink and chum salmon ([@ward_evaluating], cite more specific #?) and production has increased since the late 1970s (217 million juvenile pink salmon in PWS in 1985 versus 672 million in 2014). Though our analysis with CFEC data doesn't separate hatchery versus wild fish, as a contrast we can make coarse comparisons across regions and species to examine potential effects of hatchery fish on variability. For example, we can compare the variation in the largely hatchery pink salmon returns (earnings and weight) from Prince William Sound to the largely wild pink salmon returns from Southeast Alaska. Similarly, we can compare the variability in earnings from species dominated by hatchery returns (pink, chum) to those with less hatchery influence (Chinook, sockeye). To make these comparisons, we aggregated the earnings and landings by species and permit, then calculated 3-year rolling CVs. As a second comparison, we examined whether long-term participants in any of the fisheries had experienced a change in variability. Similar to [@kasperski_income_2013] who calculated CVs of vessels before and after significant management events, we identified individuals who fished the same permit for at least three years in the first decade of our data (1985-1995) and at least three years in the last decade of the data (2004-2014).

@@

[Rich's note: This and delta D might be tricky for pink salmon stocks, which alternate between strong even or odd years (depends on the area) and may dictate fishery participation. For these, could we, should we, use log(St/St-2)??]

@@

PWS chum and sockeye (increasing hatchery % over time) vs. Bristol Bay (all wild).

CV of variability for high value species (sockeye, Chinook) vs. low value species (pink and chum). Coho fall somewhere in the middle. In other words, how have choices of species to produce by hatcheries affected the CV of revenue for individual permits? % hatchery fish of the harvest could be used as a continuous variable in the model.

Similar to above, do life-history characteristics influence the variability of revenue? For example, sockeye and chum have substantially more complicated life-history traits (1-5 ocean years) than pink salmon (1 ocean year only). For high-volume hatcheries such as pink salmon: do annual returns affect price and revenue the following year more than for high-value fisheries?

@@

- In terms of participation, why did people drop out in 2000 in SEAK and Alaska Peninsula, but not in PWS? PWS largely seems like a boutique fishery on the Copper River -- which includes a qualtiy component (Copper River at forefront) and a timing issue relative to the rest of the Gulf of Alaska populations (buffers against market impacts). In addition, there are many indications that costs have increased over time as a result of requirements by fish buyers for fish of higher quality (refrigerated sea water, ice, increased handling time, more expensive vessels) more expensive permits, etc.

@@

- One potential exogenous driver related to management that can be ruled out as being responsible for increased specialization of salmon fishermen is changes in the timing or length of fishing seasons. Shifting seasons to later in the year may reduce catch of early run fish for example. However, we found no significant trend in either the season length or average season date for any fishery (Fig. A15). Also, management actions over time have generally tried to eliminate interception fisheries. Examples include the closure of the South Kayak Island area where sockeye bound for Cook Inlet were being harvested off the Copper River. Rich can provide some citations.

- The decisions to implement a hatchery program for a given species has economic consequences for overall revenue and the variability of that revenue to fishers and other participants. Salmon that have complex life histories in the ocean phase of their life cycle (sockeye and possibly chum salmon) may have an inherent portfolio that buffers against large annual swings in abundance compared to pink and coho salmon. Also, high-value species may experience less annual fluctuation than lower value species.

- The potential to diversify may be limited for areas with a limited number of available species or by managements actions that prohibit fishing in certain areas (e.g., PWS where seiners are generally confined to areas/times with pink and chum salmon whereas gillnet fishers are able to fish in areas with sockeye, chum, Chinook, coho, and pink salmon). Rich can provide a citation for the PWS management plan.
